---
title: "RB2OC - Data Analysis"
author: "Jamie Kleer"
date: "2025-07-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(scales)
```

# Preface  

This markdown is the basis for the bachelor's thesis done by Jamie Kleer under supervision of PD Dr. Maïwen Caudron-Herger and Prof. Dr. Benedikt Brors at DKFZ (German Cancer Research Center).  
The project consists of a data analysis performed in R concerning the effect of four proteins - TIA1, TIAL1, TPX2, and AURKA on breast and ovarian cancer patients. It will compare protein and mRNA expression with clinical data, genomic stability as well as their correlation with mutations or expression of tumor suppressor genes BRCA1, BRCA2, and TP53. The data for this study is provided by the TCGA PanCancer Atlas.  

# Opening Data

```{r}
bc_pe_raw = read.delim("data/BC_TCGA_PanCan/Protein expression z-scores (mass spectrometry by CPTAC).txt", header = TRUE,sep = "\t")  
bc_mrna_raw = read.delim("data/BC_TCGA_PanCan/mRNA expression z-scores relative to all samples (log RNA Seq V2 RSEM).txt", header = TRUE, sep = "\t")
bc_cd_raw = read.delim("data/BC_TCGA_PanCan/brca_tcga_pan_can_atlas_2018_clinical_data.tsv", 
                       header = TRUE)  
bc_cna_raw = read.delim("data/BC_TCGA_PanCan/cna.txt", header = TRUE, sep = "\t")
bc_mut_raw = read.delim("data/BC_TCGA_PanCan/mutations.txt", header = TRUE, sep = "\t") 

oc_pe_raw = read.delim("data/OC_TCGA_PanCan/Protein level z-scores (mass spectrometry by CPTAC).txt", header = TRUE,sep = "\t")  
oc_mrna_raw = read.delim("data/OC_TCGA_PanCan/mRNA expression z-scores relative to all samples (log RNA Seq V2 RSEM).txt", header = TRUE, sep = "\t")
oc_cd_raw = read.delim("data/OC_TCGA_PanCan/ov_tcga_pan_can_atlas_2018_clinical_data.tsv", header = TRUE)  
oc_cna_raw = read.delim("data/OC_TCGA_PanCan/cna.txt", header = TRUE, sep = "\t")
oc_mut_raw = read.delim("data/OC_TCGA_PanCan/mutations.txt", header = TRUE, sep = "\t")
```

# Cleaning Data

```{r}
cd_columns = c("Sample.ID", "Overall.Survival..Months.", "Aneuploidy.Score", "Mutation.Count", "MSI.MANTIS.Score", "MSIsensor.Score", "Fraction.Genome.Altered")

bc_pe = na.omit(bc_pe_raw[, -c(1, 7)])
bc_mrna = na.omit(bc_mrna_raw[, -1])
bc_cd = na.omit(bc_cd_raw[, cd_columns])
colnames(bc_cd)[c(1, 2)] = c("SAMPLE_ID", "Survival.[months]")
bc_cna = bc_cna_raw[!rowSums(bc_cna_raw == "NP"), -1]
bc_mut = bc_mut_raw[!rowSums(bc_mut_raw == "NP"), -1]

oc_pe = na.omit(oc_pe_raw[, -c(1, 6, 7)])
oc_mrna = na.omit(oc_mrna_raw[, -1])
oc_cd = na.omit(oc_cd_raw[, cd_columns])
colnames(oc_cd)[c(1, 2)] = c("SAMPLE_ID", "Survival.[months]")
oc_cna = oc_cna_raw[!rowSums(oc_cna_raw == "NP"), -1]
oc_mut = oc_mut_raw[!rowSums(oc_mut_raw == "NP"), -1]
```

# Merging Data

```{r}
bc_cd_pe = merge(bc_cd, bc_pe, by = "SAMPLE_ID")
bc_cd_pe_mut = merge(bc_cd_pe, bc_mut, by = "SAMPLE_ID")
bc_cd_mrna = merge(bc_cd, bc_mrna, by = "SAMPLE_ID")
bc_cd_mrna_mut = merge(bc_cd_mrna, bc_mut, by = "SAMPLE_ID")
bc_pe_mrna = merge(bc_pe, bc_mrna, by = "SAMPLE_ID")
bc_pe_mrna_mut = merge(bc_pe_mrna, bc_mut, by = "SAMPLE_ID")
bc_pe_cna = merge(bc_pe, bc_cna, by = "SAMPLE_ID")
bc_mrna_cna = merge(bc_mrna, bc_cna, by = "SAMPLE_ID")

oc_cd_pe = merge(oc_cd, oc_pe, by = "SAMPLE_ID")
oc_cd_pe_mut = merge(oc_cd_pe, oc_mut, by = "SAMPLE_ID")
oc_cd_mrna = merge(oc_cd, oc_mrna, by = "SAMPLE_ID")
oc_cd_mrna_mut = merge(oc_cd_mrna, oc_mut, by = "SAMPLE_ID")
oc_pe_mrna = merge(oc_pe, oc_mrna, by = "SAMPLE_ID")
oc_pe_mrna_mut = merge(oc_pe_mrna, oc_mut, by = "SAMPLE_ID")
oc_pe_cna = merge(oc_pe, oc_cna, by = "SAMPLE_ID")
oc_mrna_cna = merge(oc_mrna, oc_cna, by = "SAMPLE_ID")
```

# Plotting Data  

```{r}
genes = c("TPX2", "AURKA", "TIA1", "TIAL1", "BRCA1", "BRCA2", "TP53")
genes_pe_bc = c("TPX2", "AURKA", "TIA1", "TIAL1", "BRCA1", "TP53")
genes_pe_oc = c("TPX2", "TIA1", "TIAL1", "BRCA2", "TP53")
data_list = list(bc_cd_pe_mut, bc_cd_mrna_mut, oc_cd_pe_mut, oc_cd_mrna_mut)
cancer_list = c("BC", "BC", "OC", "OC")
expr_list = c("Protein", "mRNA", "Protein", "mRNA")
x_lim = c(-3, 3)
```

## Clinical Data vs Protein Expression  

```{r}
plot_cd_expr = function(data, genes, expr_type, cancer_type, clinic, 
                        x_lim = NULL, y_lim = NULL, use_ranks = FALSE) {
  par(mfrow = c(2, 4), mar = c(4, 4, 4, 1), oma = c(0, 0, 3, 0), pty = "s")
  for (gene in genes) {
    x = data[[paste0(gene, ".x")]]
    y = data[[clinic]]
    mut_status = data[[paste0(gene, ".y")]]
    colors = ifelse(mut_status == "WT", alpha("blue", 0.5), alpha("darkmagenta", 0.5))
    
    if (use_ranks) {
      x = rank(x)
      y = rank(y)
    }
    
    model = lm(y ~ x)
    test = cor.test(x, y, method = "pearson")
    pearson_cor = round(cor(x, y, method = "pearson"), 2)
    p_value = signif(test$p.value, 2)
    n_points = length(x)
    
    x_lab = ifelse(use_ranks, sprintf("Ranked %s Expression [z-score]", gene), 
                       sprintf("%s Expression [z-score]", gene))
    y_lab = ifelse(use_ranks, sprintf("Ranked %s", clinic),
                     sprintf("%s", gsub("\\.", " ", clinic)))
    plot(
      x = x,
      y = y,
      xlim = x_lim,
      ylim = y_lim,
      xlab = x_lab,
      ylab = y_lab,
      main = sprintf("%s: r = %s, p = %s", gene, pearson_cor, p_value),
      col = colors,
      pch = 19
    )
    abline(model, col = "red", lwd = 2)
  }
  title(
    main = sprintf("%s: %s%s Expression vs %s (n = %s)", cancer_type, 
                       if (use_ranks) "Ranked " else "",
                       expr_type, gsub("\\.", " ", clinic), n_points),
    outer = TRUE,
    cex.main = 2
    )
  plot.new()
  legend(
    "center",
    legend = c("WT", "Mutated"),
    col = c(alpha("blue", 0.5), alpha("darkmagenta", 0.5)),
    pch = 19,
    pt.cex = 1.5,
    horiz = TRUE,
    bty = "n"
  )
}
```

```{r}
plot_focus_cd = function(focus, data_list, cancer_list, expr_list, clinic, x_lim = NULL, y_lim = NULL) {
  valid_datasets = sapply(data_list, function(dataset) {
    all(c(paste0(focus, ".x"), paste0(focus, ".y")) %in% names(dataset))
  })
  par(mfrow = c(1, 4), mar = c(4, 4, 5, 1), oma = c(0, 0, 1, 0), pty = "s")
  for (i in which(valid_datasets)){
    data = data_list[[i]]
    cancer_type = cancer_list[[i]]
    expr_type = expr_list[[i]]
    
    x = data[[paste0(focus, ".x")]]
    y = data[[clinic]]
    mut_status = data[[paste0(focus, ".y")]]
    colors = ifelse(mut_status == "WT", alpha("blue", 0.5), alpha("darkmagenta", 0.5))
    
    model = lm(y ~ x)
    test = cor.test(x, y, method = "pearson")
    pearson_cor = round(cor(x, y, method = "pearson"), 2)
    p_value = signif(test$p.value, 2)
    n_points = length(x)
    
    plot(
      x = x,
      y = y,
      xlim = x_lim,
      ylim = y_lim,
      xlab = sprintf("%s Expression [z-score]", focus),
      ylab = sprintf("%s", gsub("\\.", " ", clinic)),
      main = sprintf("%s: %s (n = %s)", cancer_type, expr_type, n_points),
      mtext(sprintf("r = %s, p = %s", pearson_cor, p_value), side = 3, line = 0.5, cex = 0.8),
      col = colors,
      pch = 19
    )
    abline(model, col = "red", lwd = 2)
    legend(
      "topright",
      legend = c("WT", "Mutated"),
      col = c(alpha("blue", 0.5), alpha("darkmagenta", 0.5)),
      pch = 19,
      pt.cex = 1.5,
    )
  }
  title(
    main = sprintf("%s Expression vs %s", focus, gsub("\\.", " ", clinic)),
    outer = TRUE,
    cex.main = 2
  )
}
```

### Overall Survival Months  

```{r, fig.height = 6, fig.width = 9}
y_lim = c(0, 200)
plot_cd_expr(bc_cd_pe_mut, genes_pe_bc, "Protein", "BC", "Survival.[months]", x_lim, y_lim)
plot_cd_expr(bc_cd_mrna_mut, genes, "mRNA", "BC", "Survival.[months]", x_lim, y_lim)
plot_cd_expr(oc_cd_pe_mut, genes_pe_oc, "Protein", "OC", "Survival.[months]", x_lim, y_lim)
plot_cd_expr(oc_cd_mrna_mut, genes, "mRNA", "OC", "Survival.[months]", x_lim, y_lim)
plot_cd_expr(oc_cd_pe_mut, "TIAL1", "Protein", "OC", "Survival.[months]", x_lim, y_lim)
```

```{r, fig.height = 3, fig.width = 9}
plot_focus_cd("TPX2", data_list, cancer_list, expr_list, "Survival.[months]", x_lim, y_lim)
```

### Aneuploidy Score  

```{r, fig.height = 6, fig.width = 9}
y_lim = c(0, 30)
plot_cd_expr(bc_cd_pe_mut, genes_pe_bc, "Protein", "BC", "Aneuploidy.Score", x_lim, y_lim)
plot_cd_expr(bc_cd_mrna_mut, genes, "mRNA", "BC", "Aneuploidy.Score", x_lim, y_lim)
plot_cd_expr(oc_cd_pe_mut, genes_pe_oc, "Protein", "OC", "Aneuploidy.Score", x_lim, y_lim)
plot_cd_expr(oc_cd_mrna_mut, genes, "mRNA", "OC", "Aneuploidy.Score", x_lim, y_lim)
plot_cd_expr(oc_cd_pe_mut, c("TPX2", "TIA1", "TP53"), "Protein", "OC", "Aneuploidy.Score", x_lim, y_lim)
```
```{r, fig.height = 3, fig.width = 9}
plot_focus_cd("TPX2", data_list, cancer_list, expr_list, "Aneuploidy.Score", x_lim, y_lim)
plot_focus_cd("AURKA", data_list, cancer_list, expr_list, "Aneuploidy.Score", x_lim, y_lim)
plot_focus_cd("TIA1", data_list, cancer_list, expr_list, "Aneuploidy.Score", x_lim, y_lim)
plot_focus_cd("BRCA1", data_list, cancer_list, expr_list, "Aneuploidy.Score", x_lim, y_lim)
plot_focus_cd("BRCA2", data_list, cancer_list, expr_list, "Aneuploidy.Score", x_lim, y_lim)
```

### Mutation Count  

```{r, fig.height = 6, fig.width = 9}
y_lim = c(0, 300)
plot_cd_expr(bc_cd_pe_mut, genes_pe_bc, "Protein", "BC", "Mutation.Count", x_lim, y_lim)
plot_cd_expr(bc_cd_mrna_mut, genes, "mRNA", "BC", "Mutation.Count", x_lim, y_lim)
plot_cd_expr(oc_cd_pe_mut, genes_pe_oc, "Protein", "OC", "Mutation.Count", x_lim, y_lim)
plot_cd_expr(oc_cd_mrna_mut, genes, "mRNA", "OC", "Mutation.Count", x_lim, y_lim)
```

```{r, fig.height = 3, fig.width = 9}
plot_focus_cd("TP53", data_list, cancer_list, expr_list, "Mutation.Count", x_lim, y_lim)
```

### Fraction Genome Altered 

```{r, fig.height = 6, fig.width = 9}
y_lim = c(0, 1)
plot_cd_expr(bc_cd_pe_mut, genes_pe_bc, "Protein", "BC", "Fraction.Genome.Altered", x_lim, y_lim)
plot_cd_expr(bc_cd_mrna_mut, genes, "mRNA", "BC", "Fraction.Genome.Altered", x_lim, y_lim)
plot_cd_expr(oc_cd_pe_mut, genes_pe_oc, "Protein", "OC", "Fraction.Genome.Altered", x_lim, y_lim)
plot_cd_expr(oc_cd_mrna_mut, genes, "mRNA", "OC", "Fraction.Genome.Altered", x_lim, y_lim)
```
```{r, fig.height = 3, fig.width = 9}
plot_focus_cd("TPX2", data_list, cancer_list, expr_list, "Fraction.Genome.Altered", x_lim, y_lim)
plot_focus_cd("AURKA", data_list, cancer_list, expr_list, "Fraction.Genome.Altered", x_lim, y_lim)
plot_focus_cd("TIA1", data_list, cancer_list, expr_list, "Fraction.Genome.Altered", x_lim, y_lim)
```

### MSI MANTIS Score 

```{r, fig.height = 6, fig.width = 9}
y_lim = c(0.25, 0.55)
plot_cd_expr(bc_cd_pe_mut, genes_pe_bc, "Protein", "BC", "MSI.MANTIS.Score", x_lim, y_lim)
plot_cd_expr(bc_cd_mrna_mut, genes, "mRNA", "BC", "MSI.MANTIS.Score", x_lim, y_lim)
plot_cd_expr(oc_cd_pe_mut, genes_pe_oc, "Protein", "OC", "MSI.MANTIS.Score", x_lim, y_lim)
plot_cd_expr(oc_cd_mrna_mut, genes, "mRNA", "OC", "MSI.MANTIS.Score", x_lim, y_lim)
```

```{r, fig.height = 3, fig.width = 9}
plot_focus_cd("TIA1", data_list, cancer_list, expr_list, "MSI.MANTIS.Score", x_lim, y_lim)
```

### MSIsensor Score 

```{r, fig.height = 6, fig.width = 9}
y_lim = c(0, 10)
plot_cd_expr(bc_cd_pe_mut, genes_pe_bc, "Protein", "BC", "MSIsensor.Score", x_lim, y_lim)
plot_cd_expr(bc_cd_mrna_mut, genes, "mRNA", "BC", "MSIsensor.Score", x_lim, y_lim)
plot_cd_expr(oc_cd_pe_mut, genes_pe_oc, "Protein", "OC", "MSIsensor.Score", x_lim, y_lim)
plot_cd_expr(oc_cd_mrna_mut, genes, "mRNA", "OC", "MSIsensor.Score", x_lim, y_lim)
```

```{r, fig.height = 3, fig.width = 9}
plot_focus_cd("TPX2", data_list, cancer_list, expr_list, "MSIsensor.Score", x_lim, y_lim)
plot_focus_cd("AURKA", data_list, cancer_list, expr_list, "MSIsensor.Score", x_lim, y_lim)
plot_focus_cd("BRCA2", data_list, cancer_list, expr_list, "MSIsensor.Score", x_lim, y_lim)
```

## Expressions vs Expression 

```{r}
plot_expr_expr = function(data, focus, genes, expr_type, cancer_type, x_lim = NULL, y_lim = NULL){
  other_genes = setdiff(genes, focus)
  par(mfrow = c(2, 4), mar = c(4, 4, 4, 1), oma = c(0, 0, 3, 0), pty = "s")
  for (gene in other_genes) {
    x = data[[paste0(focus, ".x")]]
    y = data[[paste0(gene, ".x")]]
    mut_status = data[[paste0(gene, ".y")]]
    colors = ifelse(mut_status == "WT", alpha("blue", 0.5), alpha("darkmagenta", 0.5))
    
    model = lm(y ~ x)
    test = cor.test(x, y, method = "pearson")
    pearson_cor = round(cor(x, y, method = "pearson"), 2)
    p_value = signif(test$p.value, 2)
    n_points = length(x)
    
    plot(
      x = x,
      y = y,
      xlim = x_lim,
      ylim = y_lim,
      xlab = sprintf("%s Expression [z-score]", focus),
      ylab = sprintf("%s Expression [z-score]", gene),
      main = sprintf("%s: r = %s, p = %s", gene, pearson_cor, p_value),
      col = colors,
      pch = 19
    )
    abline(model, col = "red", lwd = 2)
  }
  title(
    main = sprintf("%s: %s %s Co-Expression (n = %s)", cancer_type, focus, expr_type, n_points),
    outer = TRUE,
    cex.main = 2
  )
  plot.new()
  legend(
    "center",
    legend = c("WT", "Mutated"),
    col = c(alpha("blue", 0.5), alpha("darkmagenta", 0.5)),
    pch = 19,
    pt.cex = 1.5,
    horiz = TRUE,
    bty = "n"
  )
}
```

```{r}
plot_focus_expr = function(focus, data_list, cancer_list, expr_list, gene, 
                           x_lim = NULL, y_lim = NULL) {
  valid_datasets = sapply(data_list, function(dataset) {
    all(c(paste0(focus, ".x"), paste0(focus, ".y"), 
          paste0(gene, ".x"), paste0(gene, ".y")) %in% names(dataset))
  })
  par(mfrow = c(1, 4), mar = c(4, 4, 5, 1), oma = c(0, 0, 3, 0), pty = "s")
  for (i in which(valid_datasets)) {
    data = data_list[[i]]
    cancer_type = cancer_list[[i]]
    expr_type = expr_list[[i]]
    
    x = data[[paste0(focus, ".x")]]
    y = data[[paste0(gene, ".x")]]
    mut_status = data[[paste0(gene, ".y")]]
    colors = ifelse(mut_status == "WT", alpha("blue", 0.5), alpha("darkmagenta", 0.5))
    
    model = lm(y ~ x)
    test = cor.test(x, y, method = "pearson")
    pearson_cor = round(cor(x, y, method = "pearson"), 2)
    p_value = signif(test$p.value, 2)
    n_points = length(x)
    
    plot(
      x = x,
      y = y,
      xlim = x_lim,
      ylim = y_lim,
      xlab = sprintf("%s Expression [z-score]", focus),
      ylab = sprintf("%s Expression [z-score]", gene),
      main = sprintf("%s: %s (n = %s)", cancer_type, expr_type, n_points),
      mtext(sprintf("r = %s, p = %s", pearson_cor, p_value), side = 3, line = 0.5, cex = 0.8),
      col = colors,
      pch = 19
    )
    abline(model, col = "red", lwd = 2)
    legend(
      "topright",
      legend = c("WT", "Mutated"),
      col = c(alpha("blue", 0.5), alpha("darkmagenta", 0.5)),
      pch = 19,
      pt.cex = 1.5
    )
  }
  title(
    main = sprintf("%s & %s Co-Expression", focus, gene),
    outer = TRUE,
    cex.main = 2
  )
}
```

```{r}
x_lim = c(-3, 3)
y_lim = c(-3, 3)
```

```{r, fig.height = 6, fig.width = 9}
focus = "TPX2"
plot_expr_expr(bc_cd_pe_mut, focus, genes_pe_bc, "Protein", "BC", x_lim, y_lim)
plot_expr_expr(bc_cd_mrna_mut, focus, genes, "mRNA", "BC", x_lim, y_lim)
plot_expr_expr(oc_cd_pe_mut, focus, genes_pe_oc, "Protein", "OC", x_lim, y_lim)
plot_expr_expr(oc_cd_mrna_mut, focus, genes, "mRNA", "OC", x_lim, y_lim)
```

```{r, fig.height = 3, fig.width = 9}
plot_focus_expr("AURKA", data_list, cancer_list, expr_list, "TPX2", x_lim, y_lim)
plot_focus_expr("BRCA1", data_list, cancer_list, expr_list, "TPX2", x_lim, y_lim)
plot_focus_expr("BRCA2", data_list, cancer_list, expr_list, "TPX2", x_lim, y_lim)
```

```{r, fig.height = 6, fig.width = 9}
focus = "AURKA"
plot_expr_expr(bc_cd_pe_mut, focus, genes_pe_bc, "Protein", "BC", x_lim, y_lim)
plot_expr_expr(bc_cd_mrna_mut, focus, genes, "mRNA", "BC", x_lim, y_lim)

plot_expr_expr(oc_cd_mrna_mut, focus, genes, "mRNA", "OC", x_lim, y_lim)
```

```{r, fig.height = 3, fig.width = 9}
plot_focus_expr("BRCA1", data_list, cancer_list, expr_list, "AURKA", x_lim, y_lim)
plot_focus_expr("BRCA2", data_list, cancer_list, expr_list, "AURKA", x_lim, y_lim)
```

```{r, fig.height = 6, fig.width = 9}
focus = "TIA1"
plot_expr_expr(bc_cd_pe_mut, focus, genes_pe_bc, "Protein", "BC", x_lim, y_lim)
plot_expr_expr(bc_cd_mrna_mut, focus, genes, "mRNA", "BC", x_lim, y_lim)
plot_expr_expr(oc_cd_pe_mut, focus, genes_pe_oc, "Protein", "OC", x_lim, y_lim)
plot_expr_expr(oc_cd_mrna_mut, focus, genes, "mRNA", "OC", x_lim, y_lim)
```

```{r, fig.height = 3, fig.width = 9}
plot_focus_expr("TIAL1", data_list, cancer_list, expr_list, "TIA1", x_lim, y_lim)
plot_focus_expr("BRCA1", data_list, cancer_list, expr_list, "TIA1", x_lim, y_lim)
plot_focus_expr("BRCA2", data_list, cancer_list, expr_list, "TIA1", x_lim, y_lim)
```

```{r, fig.height = 6, fig.width = 9}
focus = "TIAL1"
plot_expr_expr(bc_cd_pe_mut, focus, genes_pe_bc, "Protein", "BC", x_lim, y_lim)
plot_expr_expr(bc_cd_mrna_mut, focus, genes, "mRNA", "BC", x_lim, y_lim)
plot_expr_expr(oc_cd_pe_mut, focus, genes_pe_oc, "Protein", "OC", x_lim, y_lim)
plot_expr_expr(oc_cd_mrna_mut, focus, genes, "mRNA", "OC", x_lim, y_lim)
```

```{r, fig.height = 3, fig.width = 9}
plot_focus_expr("BRCA1", data_list, cancer_list, expr_list, "TIAL1", x_lim, y_lim)
plot_focus_expr("BRCA2", data_list, cancer_list, expr_list, "TIAL1", x_lim, y_lim)
```

```{r, fig.height = 6, fig.width = 9}
focus = "BRCA1"
plot_expr_expr(bc_cd_pe_mut, focus, genes_pe_bc, "Protein", "BC", x_lim, y_lim)
plot_expr_expr(bc_cd_mrna_mut, focus, genes, "mRNA", "BC", x_lim, y_lim)

plot_expr_expr(oc_cd_mrna_mut, focus, genes, "mRNA", "OC", x_lim, y_lim)
```

```{r, fig.height = 3, fig.width = 9}
plot_focus_expr("BRCA2", data_list, cancer_list, expr_list, "BRCA1", x_lim, y_lim)
```

```{r, fig.height = 6, fig.width = 9}
focus = "BRCA2"

plot_expr_expr(bc_cd_mrna_mut, focus, genes, "mRNA", "BC", x_lim, y_lim)
plot_expr_expr(oc_cd_pe_mut, focus, genes_pe_oc, "Protein", "OC", x_lim, y_lim)
plot_expr_expr(oc_cd_mrna_mut, focus, genes, "mRNA", "OC", x_lim, y_lim)
```
```{r, fig.height = 6, fig.width = 9}
focus = "TP53"
plot_expr_expr(bc_cd_pe_mut, focus, genes_pe_bc, "Protein", "BC", x_lim, y_lim)
plot_expr_expr(bc_cd_mrna_mut, focus, genes, "mRNA", "BC", x_lim, y_lim)
plot_expr_expr(oc_cd_pe_mut, focus, genes_pe_oc, "Protein", "OC", x_lim, y_lim)
plot_expr_expr(oc_cd_mrna_mut, focus, genes, "mRNA", "OC", x_lim, y_lim)
```


## Transcriptomics Validation

```{r}
plot_pe_mrna = function(data, genes, cancer_type, x_lim = NULL, y_lim = NULL) {
  par(mfrow = c(2, 4), mar = c(4, 4, 4, 1), oma = c(0, 0, 3, 0), pty = "s")
  for (gene in genes) {
    x = data[[paste0(gene, ".x")]]
    y = data[[paste0(gene, ".y")]]
    mut_status = data[[gene]]
    colors = ifelse(mut_status == "WT", alpha("blue", 0.5), alpha("darkmagenta", 0.5))
    
    model = lm(y ~ x)
    test = cor.test(x, y, method = "pearson")
    pearson_cor = round(cor(x, y, method = "pearson"), 2)
    p_value = signif(test$p.value, 2)
    n_points = length(x)
    
    plot(
      x = x,
      y = y,
      xlim = x_lim,
      ylim = y_lim,
      xlab = "Protein Expression [z-score]",
      ylab = "mRNA Expression [z-score]",
      main = sprintf("%s: r = %s, p = %s", gene, pearson_cor, p_value),
      col = colors,
      pch = 19
    )
    abline(model, col = "red", lwd = 2)
  }
  title(
    main = sprintf("%s: mRNA vs Protein Expression (n = %s)", cancer_type, n_points),
    outer = TRUE,
    cex.main = 2
  )
  plot.new()
  legend(
    "center",
    legend = c("WT", "Mutated"),
    col = c(alpha("blue", 0.5), alpha("darkmagenta", 0.5)),
    pch = 19,
    pt.cex = 1.5,
    horiz = TRUE,
    bty = "n"
  )
}
```

```{r, fig.height = 6, fig.width = 9}
x_lim = c(-3, 3)
y_lim = c(-3, 3)
plot_pe_mrna(bc_pe_mrna_mut, genes_pe_bc, "BC", x_lim, y_lim)
plot_pe_mrna(oc_pe_mrna_mut, genes_pe_oc, "OC", x_lim, y_lim)
```

## CNA

```{r}
boxplot_cna = function(data, genes, ref_gene, expr_type, cancer_type) {
  par(mfrow = c(2, 4), mar = c(4, 4, 4, 1), oma = c(0, 0, 3, 0), pty = "s")
  
  cna = data[[paste0(ref_gene, ".y")]]
  cna_group = factor(ifelse(cna < 0, "Loss", ifelse(cna > 0, "Gain", "Neutral")),
                     levels = c("Loss", "Neutral", "Gain"))
  
  for (gene in genes) {
    expr = data[[paste0(gene, ".x")]]
    n_points = length(expr)

    boxplot(expr ~ cna_group,
            main = gene,
            ylim = c(-3, 3),
            xlab = sprintf("CNA Status of %s", ref_gene),
            ylab = sprintf("Expression of %s", gene),
            col = c("skyblue", "gray90", "salmon"),
            border = "black",
            outline = TRUE)
  }

  title(main = sprintf("%s: %s Expression by CNA of %s (n = %s)", cancer_type, expr_type, ref_gene, n_points), outer = TRUE, cex.main = 1.5)
  par(mfrow = c(1, 1), oma = c(0,0,0,0))
}
```

```{r, fig.height = 6, fig.width = 9}
boxplot_cna(bc_mrna_cna, genes, "BRCA1", "mRNA", "BC")
boxplot_cna(oc_mrna_cna, genes, "BRCA1", "mRNA", "OC")
```

```{r, fig.height = 6, fig.width = 9}
boxplot_cna(bc_mrna_cna, genes, "BRCA2", "mRNA", "BC")
boxplot_cna(oc_mrna_cna, genes, "BRCA2", "mRNA", "OC")
```

```{r, fig.height = 6, fig.width = 9}
boxplot_cna(bc_mrna_cna, genes, "TP53", "mRNA", "BC")
boxplot_cna(oc_mrna_cna, genes, "TP53", "mRNA", "OC")
```

```{r, fig.height = 6, fig.width = 9}
boxplot_cna(bc_mrna_cna, genes, "TPX2", "mRNA", "BC")
boxplot_cna(oc_mrna_cna, genes, "TPX2", "mRNA", "OC")
```

```{r, fig.height = 6, fig.width = 9}
boxplot_cna(bc_mrna_cna, genes, "AURKA", "mRNA", "BC")
boxplot_cna(oc_mrna_cna, genes, "AURKA", "mRNA", "OC")
```

```{r, fig.height = 6, fig.width = 9}
boxplot_cna(bc_mrna_cna, genes, "TIA1", "mRNA", "BC")
boxplot_cna(oc_mrna_cna, genes, "TIA1", "mRNA", "OC")
```

```{r, fig.height = 6, fig.width = 9}
boxplot_cna(bc_mrna_cna, genes, "TIAL1", "mRNA", "BC")
boxplot_cna(oc_mrna_cna, genes, "TIAL1", "mRNA", "OC")
```
